---
title: "Introduction to `{whatifbandit}`"
bibliography: "REFERENCES.bib"
vignette: >
  %\VignetteIndexEntry{Introduction to `{whatifbandit}`}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
message: FALSE
warnings: FALSE
---

# Introduction

Traditional randomized controlled trials (RCTs) are the workhorse of causal inference in the social
sciences. Randomization of treatments provides unconfoundedness, and the fixed, equal probability of treatment assignment means causal effects can be validly estimated even with the simplest, most intuitive estimators. However, these designs come with two distinct costs:

- The number of treatments to be tested must be limited or pruned in advance by researchers to maintain the experiment's power given a fixed sample size.
- Participants continue to be assigned weaker treatments, even after incoming data reveal it may be more optimal for them to receive a different one.

In a two- or three-armed, one-shot experiment, these costs are negligible, but academics and policy experimenters frequently run trials over long periods of time in which participants' welfare is directly at stake, making these costs non-negligible.

Response-adaptive designs address these costs directly by tying assignment probabilities to the
evidence a trial has already produced rather than fixing them in advance. As the data accumulates, the
probability of assignment shifts toward treatments that appear to be working, and away from ones
that aren't. This gives researchers room to test more treatments up front, since underperforming
arms will be phased out or dropped, while participants become concentrated on the treatments that
actually work as the trial unfolds.

Although there are many ways to conduct a response-adaptive design, `{whatifbandit}` uses multi-armed bandit (MAB) algorithms, since they provide a clear framework for navigating the exploration-versus-exploitation tradeoff in treatment assignment.

The central question `{whatifbandit}` is ***What if my experiment had been run as a
bandit trial instead?*** The package name reflects this directly: *bandit* for
multi-armed bandit, and *whatif* for the counterfactual question the package
lets you explore. This vignette walks through the package's two main modes of use, its supporting
design features, and how to interpret the estimates it produces. For more details on the theory
behind Multi-Arm Bandits, please see   [The Multi-Arm Bandit Problem](bandit.qmd)

# Two Ways to Explore a Bandit Design

`{whatifbandit}` supports two distinct workflows, depending on whether you
already have trial data in hand or are planning a trial that hasn't occurred
yet.

## 1. Resimulation

If you have data from a trial that already ran as a standard RCT,
`{whatifbandit}` can resimulate that trial as though a bandit algorithm had
governed assignment instead. Using the outcomes you actually observed, the
package imputes new outcomes for participants under the counterfactual adaptive
assignment path, reconstructing a plausible "what-if" version of your
experiment. This is useful for retrospectively asking whether an adaptive design (in the same
experimental conditions) would have delivered better outcomes for participants, more precise
estimates, or both.


## 2. Simulation from Scratch

If you are designing a trial that hasn't been run yet, or doing other forms of testing, `{whatifbandit}` can
simulate a bandit trial directly from a set of population parameters you supply
(for example, assumed true success probabilities for each arm). This is useful
for:

- power analysis for a planned adaptive trial,
- comparing how different bandit algorithms would behave under your assumed
 conditions, and
- exploring design choices (batch size, exploration proportion, number of arms)
 before collecting any real data.

## Scope and Limitations

`{whatifbandit}` is currently built specifically for experiments with **binary
outcomes** (success/failure), without covariates (except for discrete blocking variables). Support for other outcome types may be added in
future releases, but any design considered in this vignette assumes a binary
response.

```{r}
#| output: FALSE
library(whatifbandit)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(forcats)
```

# Example Data: `tanf`

Examples throughout this vignette use the `tanf` dataset bundled with the
package. It contains anonymized Temporary Assistance for Needy Families (TANF) recertification
records from a field experiment which tested the impact of specific notification letters on
recertification results [@moore2022]. These new notification letters were designed to reduce the
cognitive load required to understand the instructions, and were touted as a solution to reduce the
program's notorious churn (where welfare recipients let their benefits lapse and re-apply to the
program, as opposed to recertifying their benefits) [@moore2022]. The experiment was conducted with
real TANF recipients in Washington, D.C.

```{r}
data("tanf")
head(tanf)
glimpse(tanf)
```

# Core Package Features

To customize the simulation, `{whatifbandit}` provides this core set of simulation features:

- **Two bandit algorithms**: probability matching via Thompson Sampling (TS), and
UCB1 (Upper Confidence Bound). Details on each algorithm and their implementations can be found in
[The Multi-Arm Bandit Problem](bandit.html) and the documentation.
- **Flexible assignment periods**: individual-level, batch-based, or date-based
 assignment cutoffs for bandit updates.
- **A configurable lookback window and discounting**: complete control over the number of periods
used for bandit updates, and how to weight them for the computation.
- **Perfect and imperfect information**: you can simulate the real-time trials
 where outcomes are not known immediately (`delayed_feedback`), so the
 algorithm must make decisions with incomplete information, just as you would in the field.
- **Block-randomized and cluster-randomized designs**
- **Control augmentation and random assignment proportions**: these exploration parameters enforce a
minimum level of continued exploration for the algorithms, allowing for valid treatment effect estimation.
- **Joint hypothesis testing** (***Experimental***): bootstrap or randomization inference [@offer-westort2021] to test
 whether *any* treatment effect exists across arms.
- **pairwise contrasts**: pairwise contrasts can be automatically computed versus the control, versus the current best arm,
 or all pairwise comparisons among treatments.

The package is also built to scale effectively. When conducting multiple
replications of the same design, parallel processing can be used via
[future](https://future.futureverse.org/). [data.table](https://rdatatable.gitlab.io/data.table/) is
also supported for potentially large datasets.

I encourage a look at the documentation for `mab_from_rct()`, `simulate_mab()`, and `joint_test()`
for a more detailed look at all these features and their defaults.

# Resimulating an RCT as a Bandit Trial

## A Single Resimulation

To resimulate an RCT as a bandit trial, use `mab_from_rct()`. It requires a formula relating the
outcome to the treatment, with any potential blocking or clustering included in `block()` and
`cluster()` wrappers, and how the cutoff points are determined.


```{r}
set.seed(543645)
sim <- mab_from_rct(
  success ~ condition,
  data = tanf,
  algorithm = "ucb1",
  period_method = "batch",
  period_length = 1000,
  whole_experiment = TRUE
)
```

Here, `period_method = "batch"` with `period_length = 1000` tells the algorithm
to recompute assignment probabilities every 1,000 participants.
`whole_experiment = TRUE` means the probabilities used to impute new outcomes are estimated from
the entire dataset (instead of the data received up to that point).

## Blocking and Date-Based Periods with Delayed Feedback

The original experiment in @moore2022 randomized within blocks defined by the closest service
center. There is also a natural delay from when notification letters were assigned (and sent)
compared to when researchers received the recertification outcome. The example below accounts for these features to make the
experiment a more realistic resimulation.

```{r}
future::plan("multisession", workers = 2)
set.seed(53254339)
simulations <- mab_from_rct(
  success ~ condition + block(service_center),
  data = tanf,
  algorithm = "thompson",
  period_method = "date",
  time_unit = "month",
  period_length = 1,
  delayed_feedback = TRUE,
  assignment_date_col = letter_sent_date,
  success_date_col = date_of_recert,
  date_col = appt_date,
  month_col = recert_month,
  whole_experiment = FALSE,
  r = 100,
  keep_data = TRUE,
  seed = TRUE
)
future::plan("sequential")
```

Some things to highlight are:

- `block(service_center)` inside the formula tells `{whatifbandit}` to treat
 `service_center` as a blocking variable, so randomization and outcome
 imputation respect the original block structure.
- `period_method = "date"` together with `time_unit = "month"` and
 `period_length = 1` recomputes assignment probabilities once per calendar
month, rather than after a fixed number of participants. This mimics the assignment schedule of the
original experiment.
- `delayed_feedback = TRUE`, along with the `assignment_date_col`, `success_date_col`, `date_col`,
and `month_col` arguments, tells the algorithm which outcomes were actually observable by the time each new assignment decision was made, capturing the time lag.
- ` r = 100` requests 100 independent resimulations of the trial since any single resimulation is just one draw of
 assignments and imputations.
- `keep_data = TRUE` forces the `new_data` object (explained below) to be saved for each iteration. By default, they are not saved to limit the memory footprint of the final output.
- Running `future::plan("multisession")` before the call means the 100 resimulations are conducted in
parallel across multiple workers. `seed` is an argument passed to `future::future()` for parallelization, setting it to `TRUE`
ensures that, given the same seed in the previous `set.seed()`, the same results will be returned
(essentially normal R behavior for RNG with seeds). Users have complete control over the
parallelization, and can take advantage of `...` to pass arguments to `furrr::furrr_options()`
which control the parallel setup.

# Simulating a New Adaptive Trial from Scratch

When no trial data exists yet, `simulate_mab()` lets you specify assumed true
success probabilities for each arm and simulate how a bandit algorithm would
behave under those assumptions.

```{r}
p <- matrix(
  c(0.20, 0.35, 0.5, 0.27, 0.33),
  ncol = 1,
  dimnames = list(
    c("control", "T1", "T2", "T3", "T4"),
    NULL
  )
)
print(p)

set.seed(123)
sim_from_scratch <- simulate_mab(
  n = 2000,
  t = 20,
  p = p,
  algorithm = "thompson",
  random_assign_prop = 0.3,
  contrasts = "all"
)
```

In this example:

- `p` encodes the researcher's assumed true success probability for each arm.
- `n = 2000` sets the total number of simulated participants, and `t = 20` sets
 the number of periods over which assignment probabilities are updated.
- `algorithm = "thompson"` selects Thompson sampling as
 the bandit algorithm, as opposed to `"ucb1"`.
- `random_assign_prop = 0.3` enforces that 30% of assignment probability mass is
 held out for random (non-adaptive) assignment across arms in every period,
 guaranteeing a minimum level of continued exploration rather than letting the
 algorithm fully commit to the arm it currently believes is best.
- `contrasts = "all"` will compute all $\binom{5}{2}$ pairwise effect contrasts between the treatment arms.

# Understanding The Output

Both `mab_from_rct()` and `simulage_mab()` return an S3 `.mab` class with the same fields.
Additionally, each object has a second class identifier, marking which function produced it and
whether multiple trials were run: `single_rct_mab`, `multi_rct_mab`, `single_param_mab`, and
`multi_param_mab`. Using the computed examples above:

```{r}
class(sim_from_scratch)
class(sim)
class(simulations)
```

This class structure at the moment has no implemented generics, but plans in the future are to write
generics which expedite common analyses on the final output objects.

Regardless of function or number of replications conducted, every `.mab` object contains the same
fields.

First, the `new_data` from the resulting MAB procedure is stored, allowing users to interact
with the results directly.^[By default when multiple simulations are conducted `new_data` is not
returned to limit the memory required for the output object, but can be requested by setting
`keep_data = TRUE`.]
```{r}
head(sim$new_data)
```

Second, `bandit` contains the diagnostic information of the bandit procedure, which can be used to
examine the algorithm's behavior (`statistic`) and the resulting assignment_probabilities
(`assignment_prob`) over the course of the experiment. The final assignment counts to each treatment
are included as well (`assignment_quant`). Below is a glimpse of each object, along with an example
plot of the assignment probabilities over time for @moore2022 MAB replication.

```{r}
head(sim_from_scratch$bandit$statistic)
head(sim_from_scratch$bandit$assignment_prob)
sim_from_scratch$bandit$assignment_quant

# Assignment probabilities over time
simulations$bandit$assignment_prob |>
  pivot_longer(
    cols = c(no_letter, open_appt, specific_appt),
    values_to = "p",
    names_to = "mab_condition"
  ) |>
  summarize(
    avg_p = mean(p),
    sd = sd(p),
    .by = c(mab_condition, period_number)
  ) |>
  mutate(
    lo = avg_p - 2 * sd,
    hi = avg_p + 2 * sd,
    mab_condition = str_to_title(str_replace(mab_condition, "_", " "))
  ) |>
  ggplot(aes(
    x = period_number,
    y = avg_p,
    color = mab_condition,
    fill = mab_condition
  )) +
  geom_ribbon(aes(ymax = hi, ymin = lo), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.8) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Period",
    y = "Probability of Assignment",
    title = "Assignment Probability to Treatment
Arms Over Time",
    subtitle = "With 2 Standard Deviation Uncertainty Ribbon",
    color = "Treatment Arm",
    fill = "Treatment Arm"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(face = "italic")
  )
```

Third, `means` and `contrasts` contain the final conditional expectation estimates for each treatment
arm (`means`) and treatment effect estimates (`contrasts`) for the requested configuration, along
with their standard errors.
Treatment effects can be requested via the `contrasts` argument for the following configurations:

- `"best"`: effect relative to the best treatment arm, as determined by the maximum UCB1 value or
TS probabilities.
- `"control"`: effect relative to the labelled control arm.
- `"both"`: both of the above
- `"all"`: all $\binom{n}{k}$ pairwise differences among arms

Estimates are reported for the AW-AIPW estimator [@hadad2021a], the IPW estimator
[@offer-westort2021], and a traditional OLS estimator. More details about the properties of these
estimators are available in [Causal Inference With Bandits](estimate.html), but a quick overview:

- AW-AWIPW is unbiased and asymptotically normal [@hadad2021a], so it can be used for hypothesis
tests with the standard normal distribution.
- IPW is unbiased with a non-normal asymptotic sampling distribution, but @offer-westort2021
that in extremely large samples t-tests may still provide proper coverage.
- OLS is biased and completely invalid to use for inference, but is provided for comparison
nonetheless.

Below is a glimpse of each object, along with examples of a confidence interval plot for the AW-AIPW
mean estimates, and a p-value computation for the AW-AIPW treatment effects.

```{r}
head(sim_from_scratch$means)
# Plotting 95% CI using Normal distribution for AW-AIPW estimates
sim_from_scratch$means |>
  filter(estimator == "AW-AIPW") |>
  mutate(
    lo = mean + qnorm(0.025) * se,
    hi = mean + qnorm(0.975) * se,
    mab_condition = str_to_title(mab_condition),
    mab_condition = fct_reorder(mab_condition, mean)
  ) |>
  ggplot(aes(x = mean, y = mab_condition)) +
  geom_errorbar(
    aes(xmin = lo, xmax = hi),
    orientation = "y",
    width = 0.15,
    color = "grey40",
    linewidth = 0.6
  ) +
  geom_point(size = 2.5, color = "#2c3e50") +
  theme_minimal(base_size = 12) +
  labs(
    y = NULL,
    x = "Probability of Success",
    title = "Estimated Probability of Success By Arm",
    subtitle = "AW-AIPW Point Estimates with 95% Confidence Interval"
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(face = "italic")
  )
# Simple test of no treatment effect between each pair
sim_from_scratch$contrasts |>
  filter(estimator == "AW-AIPW") |>
  mutate(
    z_stat = est / se,
    p_value = pnorm(abs(z_stat), lower.tail = FALSE) * 2
  ) |>
  select(arm1, arm2, est, se, z_stat, p_value)
```

The other objects returned are the F-statistics from the IPW and OLS regressions (`f-stats`), the
`lm_robust` objects from the IPW and OLS regressions (`models`),^[Only provided if clustering is
used so that arbitrary contrasts can be estimated later with the appropriate CR2 covariance matrix
estimator using `clubSandwich::linear_contrast()`. By default, with multiple simulations, these are
not returned since each is large memory-wise but can be requested by setting `keep_models = TRUE`.]
and a `config` object containing the arguments and original `call` object for easy replication since
there are so many customizable settings.

# Joint Hypothesis Tests

`{whatifbandit}` also implements Multi-Arm-Bandit-aware joint hypothesis tests either via a randomization inference procedure in
@offer-westort2021 or a bootstrap test, testing whether any treatment effect exists across arms.
Both tests are experimental, unverified, and noted to have low power, but are included for
completeness nonetheless. Both tests use the F-statistic from the IPW regression as its test
statistic, and simulate its sampling distribution under the appropriate $H_0$ to compute a p-value. The logic behind each test is further explained in [Causal Inference With Bandits](estimate.html)

```{r}
boot <- joint_test(sim_from_scratch, method = "bootstrap", r = 100)
rand <- joint_test(sim_from_scratch, method = "randomization", r = 100)
```

Both return a list with the observed F-statistic (`F-statistic`), the null distribution generated
under $H_0$ (`null_distribution`), the resulting `p_value` (The proportion of null F-statistics
greater than the observed one), the `method` used, the number of simulations to perform the test (`
r`), and the number of those simulations with a non-`NA` F-statistic (`effective_r`; Some iterations
can produce extremely skewed treatment assignment, resulting in an undefined F-statistic)

```{r}
boot$f_statistic
boot$p_value
boot$method
```

# Adaptive Vs. Static

[Section in Progress]


# Adaptive Vs. Static

[Section in Progress]



# Future Plans
Plans for this package are ambitious and expansive:

- Support for Contextual Bandits (covariate-aware adaptation).
- Continuous outcomes.
- Post-simulation helper functions to accelerate analysis.
- Fixed confidence settings and early stopping for fixed-budget settings.
- Adaptive treatment pool, as in switching treatments in and out of the potential assignment pool
over time. This mimics an infinitely running experiment where new treatments are added in real time.
- Exporting modular internal functions for more fine-tuned user control, and to help others run
their own adaptive experiments in the field.

# Getting Help and Contributions

For more complete details on individual function arguments and behavior, consult
the full package documentation (`?mab_from_rct`, `?simulate_mab`, `?joint_test`). If you want to get an overview of Multi-Arm Bandits or the mechanics of causal
inference with adaptive experiments, take a look at the other vignettes.

- [The Multi-Arm Bandit Problem](bandit.html)
- [Causal Inference With Bandits](estimate.html)

If you have specific questions about the package, feel free to reach out by
email at <no9857a@american.edu>. If you encounter a bug, please open an issue on
[GitHub](https://github.com/Noch05/whatifbandit/issues) with a reproducible
example.

# References

:::{#refs}
:::